---
title: "Codebook for Project Assignment in Getting and Cleaning Data"
author: "YXYJack"
date: "January 29, 2016"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```
```{r include=FALSE}
source('run_analysis.R')
```

#Study Design
***
This activity analyzes the UCI HAR Dataset.zip (considered the raw data here) which represents data collected from the accelerometers from the Samsung Galaxy S smartphone carried by a group of 30 volunteers. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. 

The analysis described in this codebook calculates the means of the variables in the raw data set grouped by person and activity.  The means have a possible range of [-1,1]; the units are either standard gravity units or radians/second, depending on the particular field.  The means are stored in a tidy data-set "TidyDataSetForCleaningProject.txt" 

The complete list of variables of each feature and summary information is available in UCI HAR Dataset.zip/features.txt and UCI HAR Dataset.zip/features_info.txt  See the codebook section for instruction on obtaining the raw data.

#Codebook
***
For convenience, the codebook is divided into the following subsections:

1. variables in TidyDataSetForCleaningProject.txt
2. variables in the data set not contained in the tidy data
3. description of transformation creating "TidyDataSetForCleaningProject.txt"

##Variables in TidyDataSetForCleaningProject.txt

Consists of the following column names and descriptions:

1. subject: Identity of volunteer.  Range 1-30
2. activityName: activity performed when measurements were collected.  Range:{WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING} 
3. columns 3-81: feature names as described in UCI HAR Dataset.zip/features.txt
  + *acc* fields: The acceleration signal from the smartphone accelerometer in standard gravity units 'g'.  Mean of values from raw data-set.  Raw data-set values normalized and bounded to [-1,1] 
  + *gyro* fields: The angular velocity vector measured by the gyroscope for each window sample. The units are radians/second.  Mean of values from raw data-set.  Raw data-set values normalized and bounded to [-1,1]

## Variables in the data set not contained in the tidy data

The following UCI HAR Dataset.zip files are directly or indirectly represented in "TidyDataSetForCleaningProject.txt" **all others are excluded**

* y_train.txt
* X_train.txt
* subject_train.txt
* y_test.txt
* X_test.txt
* subject_test.txt
* features.txt
* activity_labels.txt

The complete list of variables of each feature and summary information is available in UCI HAR Dataset.zip/features.txt and UCI HAR Dataset.zip/features_info.txt.

A full description of the raw data is available at the site where the data was obtained:
http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones

The raw data itself is available here:
https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip

##Description of transformation creating "TidyDataSetForCleaningProject.txt"
**refer to the accompanying R script: "run_analysis.R"**

The transformation consists of the following major functions:

1. Retrieving, loading and assembling one data-set from UCI HAR Dataset.zip data
2. Manipulating data
3. Writing tidy data to output file

###Retrieving and loading UCI HAR Datast.zip data

* Retrieves a copy of "UCI HAR Dataset.zip", if a local copy doesn't already exist
* Loads seven different /*.txt data-sets from the raw data using `unz': y_train,y_test,X_test,X_train,subject_test,subject_train,features
* successively merges the different data-sets in pairs using `rbind` and `cbind` until a single data-set results.Result stored as dt.  
  + Example follows:
```{r}
head(dt[,1:4],n=3)
```

###Manipulating data

* Extract only the measurements on the mean and standard deviation for each field, using `grepl` on "mean" and/or "std".  Result stored in dtExtract
* Apply descriptive activity names by loading activity_labels.txt from the UCI HAR .zip and applying to dtExtract.  Manipulate (drop, reorder) certain columns for readability.  Result stored in dtExtractNamed
* Create a different/unique tidy data-set by calculating the mean of each variable in dtExtractNamed.  Result stored in dtMean. 
* Change title of field variable names to reflect that they are now a 'mean of means'
  + Sample of dtMean follows:
```{r}
head(dtMean[,1:5],n=3)
```
###Writing tidy data to output file
*Checks if "TidyDataSetForCleaningProject.txt" already exists in the working directory, and if not writes it as a .txt file.